---
title: "Sanket Project"
author: "Sanket"
date: "`r Sys.Date()`"
output: html_document
---


```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))

#install.packages
library(readxl)
library(knitr)
library(TTR)
library(dplyr)
library(ggplot2)
library(forecast)
library(tidyverse)
library(tseries)

# Load the Excel file
data  <- read.csv("C:/Users/Sanket Khamkar/Downloads/Data Dictionary_Sanket.csv")
names(data)
knitr::kable(data)
head(data)

# Function to convert 'Year_Quarter' to start date for each quarter
convert_quarter_to_date <- function(quarter) {
  if (is.na(quarter)) {
    return(NA)  # Return NA if quarter is missing
  }
  year <- as.numeric(substr(quarter, 1, 4))
  q <- substr(quarter, 6, 7)
  
  if (q == "Q1") {
    return(as.Date(paste0(year, "-01-01")))
  } else if (q == "Q2") {
    return(as.Date(paste0(year, "-04-01")))
  } else if (q == "Q3") {
    return(as.Date(paste0(year, "-07-01")))
  } else if (q == "Q4") {
    return(as.Date(paste0(year, "-10-01")))
  } else {
    return(NA)  # Return NA if the quarter is not in a recognized format
  }
}

# Apply the conversion to the 'Year Quarter' column
data$Date <- sapply(data$`Year_Quarter`, convert_quarter_to_date)

# Convert the 'Rate' column to numeric and clean up the data
data$FertilityRate <- as.numeric(data$FertilityRate)
data <- data[!is.na(data$Date) & !is.na(data$FertilityRate), ]
data$MortalityRate <- as.numeric(data$MortalityRate)
data <- data[!is.na(data$Date) & !is.na(data$MortalityRate), ]
# Ensure Date column is a factor (or convert as needed)
data$Date <- as.factor(data$Date)

```

```{r}
# Filter for 'All races and origins' and set time series
data_filtered <- subset(data, Group == "All races and origins")
ts_data <- ts(data_filtered$FertilityRate, start = c(2022, 1), frequency = 4)

# Plot the time series
plot(ts_data , main = "Fertility Rate Time Series", ylab = "Fertility Rate", xlab = "Time")

```


```{r}
#Mean forecast
mean_model <- meanf(ts_data, h = 12)

# Plot the forecast
autoplot(mean_model) + ggtitle("Mean Forecast for Fertility Rate")

forecast(mean_model)

# Residual Analysis
checkresiduals(mean_model)
```
Uses the average of past data as the forecast. Best for stationary data without trends or seasonality. Residuals should be random, indicating that the mean is a suitable representation.


```{r}
#Naive forecast
naive_model <- naive(ts_data, h = 12)

# Plot the forecast
autoplot(naive_model) + ggtitle("Naive Forecast for Fertility Rate")

forecast(naive_model)

# Residual Analysis
checkresiduals(naive_model)

```
Suitable for short-term forecasts where the most recent observation is the best predictor. Ideal for data without trends or seasonality. Random residuals indicate a good fit.
```{r}
#Seasonal Naive forecast
snaive_forecast <- snaive(ts_data,  h = 12)
# Plot the forecast
autoplot(snaive_forecast) + ggtitle("Seasnoal Naive Forecast for Fertility Rate")

forecast(snaive_forecast)

# Residual Analysis
checkresiduals(snaive_forecast)
```
Suitable for data with strong seasonal patterns. It uses the value from the same season in the previous cycle. Residuals should be random if seasonality is captured effectively.

```{r}

#Random walk forecast
rwf_forecast <- rwf(ts_data,h = 12)
rwf_forecast <- rwf(ts_data,h = 12, drift=TRUE)
autoplot(rwf_forecast) + ggtitle("Random Walk Forecast for Fertility Rate")

forecast(rwf_forecast)

# Residual Analysis
checkresiduals(rwf_forecast)
```
Useful for data that has a persistent random walk behavior, sometimes with a drift representing a trend. Residuals should be random if the drift captures the trend accurately.

```{r}
# Simple Exponential Smoothing
ses_model <- ses(ts_data, h = 12)

# Plot the forecast
autoplot(ses_model) + ggtitle("SES Forecast for Fertility Rate")

forecast(ses_model)
# Residual Analysis
checkresiduals(ses_model)

```
Works well for level data without trend or seasonality. Residuals should resemble white noise if the model fits correctly.

```{r}
# Fit an ETS model
ets_model <- ets(ts_data)

# Forecast the next 12 months
ets_forecast <- forecast(ets_model, h = 12)

# Plot the forecast
autoplot(ets_forecast) + ggtitle("ETS Forecast for Fertility Rate")

forecast(ets_forecast)
# Residual Analysis
checkresiduals(ets_forecast)

```

Captures level, trend, and seasonality effectively. Residuals should be random and without autocorrelation to indicate a good model fit.

```{r}
#Moving Average forecast
MA5_forecast <- ma(ts_data,order=5)
# Plot the forecast
autoplot(MA5_forecast) + ggtitle("Moving Average5 Forecast for Fertility Rate")


forecast(MA5_forecast)
# Residual Analysis
checkresiduals(MA5_forecast)

#Moving Average forecast
MA9_forecast <- ma(ts_data,order=9)
# Plot the forecast
autoplot(MA9_forecast) + ggtitle("Moving Average9 Forecast for Fertility Rate")
# Residual Analysis
forecast(MA9_forecast)
checkresiduals(MA9_forecast)

```
Good for smoothing and reducing noise. Does not effectively capture trend or seasonality. Significant residual lags suggest it may not fit dynamic time series with trends.

```{r}
##Decomposition
decomp <- decompose(ts_data, type = "additive")  
# Plot the Decomposition
autoplot(decomp) + ggtitle("Decomposition Addictive Forecast for Fertility Rate")


residuals <- decomp$random

residuals_clean <- na.omit(residuals)

# Plot the ACF of the cleaned residuals
acf(residuals_clean, main = "ACF of Residuals After Decomposition (NA Removed)")

# Plot the residuals
autoplot(residuals, main = "Residuals After Decomposition") + ggtitle("Residual Component of Decomposition")

forecast(residuals_clean)

# Histogram of residuals
hist(residuals, main = "Histogram of Residuals", xlab = "Residuals", breaks = 15)

# Ljung-Box test for residuals
Box.test(residuals, lag = 10, type = "Ljung-Box")
```
Separates data into trend, seasonal, and residual components. Residual patterns indicate how well the model captures these components; randomness means effective decomposition.

```{r}
# Decompose the time series
stl_decomposition <- stl(ts_data, s.window = "periodic")

# Plot the decomposition
autoplot(stl_decomposition) + ggtitle("STL Decomposition of Fertility Rate")

# Forecast the next 12 months
stl_forecast <- forecast(stl_decomposition, h = 12)

# Plot the forecast
autoplot(stl_forecast) + ggtitle("STL Forecast for Fertility Rate")

forecast(stl_forecast)

# Residual Analysis
checkresiduals(stl_forecast)
```
Handles non-linear seasonality and trend effectively. Residuals should be random if both trend and seasonal components are adequately captured.

```{r}
# Holt-Winters Seasonal Model
hw_model <- hw(ts_data, seasonal = "additive", h = 12)

# Plot the forecast
autoplot(hw_model) + ggtitle("Holt-Winters Forecast for Fertility")

forecast(stl_forecast)
# Residual Analysis
checkresiduals(hw_model)
```
Captures both trend and seasonality, making it suitable for data with both components. Random residuals indicate a successful capture of underlying dynamics.

```{r}


# Fit an ARIMA model
arima_model <- auto.arima(ts_data)

# Forecast the next 12 months
arima_forecast <- forecast(arima_model, h = 12)

# Plot the forecast
autoplot(arima_forecast) + ggtitle("ARIMA Forecast for Fertility")

forecast(arima_forecast)
# Residual Analysis
checkresiduals(arima_model)

```
Suitable for data with auto-regressive components, differencing to make data stationary, and moving average terms. Residuals should resemble white noise without autocorrelation, indicating that the ARIMA model has properly fit the data.

```{r}
#Regression Analysis

plot(data[,-1])  # Remove non-numeric columns, if necessary
# Select only numeric columns for the pairs plot
numeric_data <- data[sapply(data, is.numeric)]


str(numeric_data)

# Plot pairs only if there are at least two numeric columns
if (ncol(numeric_data) > 1) {
    pairs(numeric_data)
} else {
    print("Not enough numeric columns for pairwise plot")
}

# Summary and correlation
summary(data)
cor(numeric_data)

# Histograms and Scatter Plots
hist(data$MortalityRate, main="Histogram of Mortality Rate", xlab="Mortality Rate")
plot(MortalityRate ~ FertilityRate, data=data)
plot(jitter(MortalityRate) ~ jitter(FertilityRate), data=data)

# Fit linear regression model
fit <- lm(MortalityRate ~ FertilityRate, data=data)
print(fit)

plot(fit)
summary(fit)

anova(fit)

# Residual analysis
res <- residuals(fit)
plot(jitter(res) ~ jitter(data$FertilityRate), main="Residuals vs Fertility Rate")
abline(h=0, col="red")
hist(res, main = "Residuals of the Model", xlab = "Residuals")

# Predicted value for the first data point
fitted(fit)[1]
```

To explain insights at the residuals, we will see differences between actual mortality rates and what our model predicted.

Residuals vs. Fitted Values Plot: Here, we’re looking for a random scatter of points around zero. If we see that randomness, it means our model is appropriate. But if we spot patterns, it might mean our model is missing something important.

Histogram of Residuals: This histogram shows the spread of prediction errors. Ideally, it looks like a bell curve centered at zero, which would indicate that most of our predictions are close to the actual values. If it’s skewed or off-center, it could mean our model is biased or not capturing some factors correctly.

Q-Q Plot: This plot checks if the residuals follow a normal distribution. If the points align closely with a straight line, it confirms that our errors are distributed as expected, reinforcing the model’s reliability. If not, it might hint at some model issues or the need for additional predictors.

Summary for Regression:
If the residuals look random and normally distributed, it shows that fertility rate is a good predictor for mortality rate in our data.
If we see patterns in residuals or deviations from normality, it might be worth revisiting the model—possibly by adding more factors, transforming some data, or even trying a different type of model to capture the relationship more accurately.

```{r}
# Accuracy checked
# Pick an accuracy measure, compare your models, and state the best model based on the accuracy comparison

accuracy(mean_model)
accuracy(naive_model)
accuracy(rwf_forecast)
accuracy(snaive_forecast)
accuracy(ses_model)
accuracy(ets_forecast)
accuracy(stl_forecast)
accuracy(hw_model)
accuracy(arima_model)
```


